---
title: "Attestations, signed bundles and tamper-evident chains"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Attestations, signed bundles and tamper-evident chains}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(rmoriebricklayer)
```

The *Building reproducible data capsules* vignette shows how a capsule pins data and
describes it in a manifest. This one covers the three ways of putting a signature on that
work so a reader can check it: an **attestation** of a manifest, a **bundle** of a whole
capsule directory, and a **chain** that makes a sequence of manifests tamper-evident.

## Attesting a manifest

An attestation binds a signature to a manifest *and* to what the manifest covers, with an
optional note from the person signing:

```{r attest}
m <- make_manifest(list(dataset = "otis", rows = 1200L), environment = FALSE)
key <- fips_keygen("ML-DSA-65")
att <- capsule_attest(m, key, note = "counts as published")

res <- capsule_check_attestation(att, m)
res$ok
res$checks
```

A verifier needs the attestation and the manifest, nothing else. Any change to the manifest
breaks it, and so does presenting a different key:

```{r attest-break}
m2 <- m
m2$meta$rows <- 1201L
capsule_check_attestation(att, m2)$ok

capsule_check_attestation(att, m, key_expected = fips_keygen("ML-DSA-65")$public)$ok
```

`key_expected` is how a reader pins the signer: the attestation carries the public key that
made it, but a reader who knows which key to expect should say so, or any key would do.

## Bundling a capsule

A bundle is one signed artifact covering a directory: every file is hashed (hidden files
included), the hashes and the manifest are signed, and the bundle verifies against the
directory later:

```{r bundle}
dir <- tempfile()
dir.create(dir)
write.csv(data.frame(x = 1:3), file.path(dir, "data.csv"), row.names = FALSE)
m <- make_manifest(list(dataset = "demo"), environment = FALSE)
key <- fips_keygen("ML-DSA-44")

b <- capsule_bundle(dir, m, key, note = "as published")
capsule_bundle_verify(attr(b, "path"), dir, manifest = m)$ok
```

Touch a byte of the data and the bundle no longer holds; the result names the file:

```{r bundle-break}
write.csv(data.frame(x = 1:4), file.path(dir, "data.csv"), row.names = FALSE)
v <- capsule_bundle_verify(attr(b, "path"), dir, manifest = m)
v$ok
```

```{r bundle-read}
capsule_bundle_read(attr(b, "path"))$attestation$note
unlink(dir, recursive = TRUE)
```

## A chain of manifests

A per-manifest signature proves each manifest is intact but says nothing about the
*sequence*: a run could be deleted from the middle of a series, or two swapped, and every
remaining signature would still verify. A hash chain links each entry to the one before it,
so deletion, insertion and reordering all break it:

```{r chain}
ch <- chain_new()
ch <- chain_append(ch, "manifest for run 1", label = "run-1")
ch <- chain_append(ch, "manifest for run 2", label = "run-2")
ch <- chain_append(ch, "manifest for run 3", label = "run-3")
chain_verify(ch)$valid
```

Editing an entry breaks the chain and names where:

```{r chain-edit}
edited <- ch
edited$entries[[2]]$digest <- core_sha256("something else")
chain_verify(edited)$valid
chain_verify(edited)$broken_at
```

Deleting one breaks it too, which a per-manifest digest would never catch:

```{r chain-drop}
dropped <- ch
dropped$entries[[2]] <- NULL
chain_verify(dropped)$valid
```

### Sealing the chain

The seal covers the links *and the length* together, so one signature over the seal fixes
the whole history. Truncating the chain still seals, but to a different value:

```{r seal}
key <- fips_keygen("ML-DSA-44")
sig <- capsule_sign(chain_seal(ch), key)
capsule_verify(chain_seal(ch), sig, fips_public_key(key))

truncated <- ch
truncated$entries[[3]] <- NULL
capsule_verify(chain_seal(truncated), sig, fips_public_key(key))
```

A chain whose links disagree has no seal to present at all:

```{r seal-none}
chain_seal(dropped)
```

## Which to use

* One manifest, one signer: `capsule_attest()`.
* A directory of files that must travel together: `capsule_bundle()`.
* A series of runs, releases or data refreshes where order and completeness matter:
  `chain_append()` after each, sign `chain_seal()` at each release.

All three accept any signing key the package knows: ML-DSA and SLH-DSA (`fips_keygen()`)
or XMSS (`pqc_keygen()`, with the key state carried forward). A shared secret from
`derive_key()` is for keyed digests (`core_hmac_sha256()`), not for attestations: an
attestation names a public key, and a shared secret has none.
