---
title: "Hawkes processes: self-exciting event series"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Hawkes processes: self-exciting event series}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 3.5)
library(rmoriebricklayer)
```

Many administrative event series cluster: one incident raises the chance of another soon
after. A **Hawkes process** (Hawkes 1971) models this directly. Its conditional intensity,
the instantaneous event rate given the past, is

$$\lambda(t) = \mu + \sum_{t_i < t} \phi(t - t_i),$$

a baseline rate $\mu$ plus a kernel $\phi$ that each past event adds and that decays with
the time since it. The integral of the kernel, $n = \int_0^\infty \phi(s)\,ds$, is the
**branching ratio**: the expected number of events each event directly triggers. When
$n < 1$ the process is stationary; $n$ is the fraction of events that are "children" of
earlier events rather than arrivals from the baseline, which is often the quantity of
substantive interest.

## Simulating a Hawkes process

Ogata's thinning algorithm simulates the process exactly: propose events from an upper
bound on the intensity and keep each with probability $\lambda(t)/\lambda_{\max}$. Here is
an exponential kernel $\phi(s) = \alpha \beta e^{-\beta s}$, whose branching ratio is
$\alpha$:

```{r simulate}
simulate_hawkes <- function(mu, alpha, beta, horizon, seed = 1) {
  set.seed(seed)
  t <- 0
  events <- numeric(0)
  while (t < horizon) {
    lambda_max <- mu + alpha * beta * sum(exp(-beta * (t - events)))
    t <- t + stats::rexp(1, lambda_max)
    if (t >= horizon) break
    lambda_t <- mu + alpha * beta * sum(exp(-beta * (t - events)))
    if (stats::runif(1) <= lambda_t / lambda_max) events <- c(events, t)
  }
  events
}
times <- simulate_hawkes(mu = 1, alpha = 0.5, beta = 2, horizon = 300)
length(times)
```

## Fitting

`core_hawkes_fit()` maximises the exact log-likelihood with an analytic gradient in C
(projected BFGS, with the parameters bounded). `theta` holds the baseline, the branching
ratio and the kernel parameter:

```{r fit}
fit <- core_hawkes_fit(times, horizon = 300, kernel = "exponential")
fit$theta
fit$branching_ratio
```

The branching ratio is recovered close to the 0.5 the data were simulated with, and the
baseline close to 1. Four kernels are available: `exponential` (memoryless decay),
`weibull` and `gamma` (a delayed peak), `lomax` (a heavy tail: influence that fades slowly).
Compare them by information criterion:

```{r kernels}
sapply(c("exponential", "weibull", "lomax"), function(k) {
  core_hawkes_fit(times, 300, kernel = k)$aic
})
```

A **sinusoidal baseline** lets the background rate follow a cycle (a weekly or seasonal
pattern) instead of staying constant:

```{r sinus}
core_hawkes_fit(times, 300, kernel = "exponential", baseline = "sinusoidal")$aic
```

## Checking the fit: time-rescaling residuals

If the fitted intensity is right, the compensator $\Lambda(t_i) = \int_0^{t_i}\lambda(s)\,ds$
turns the event times into a unit-rate Poisson process (the time-rescaling theorem; Brown et
al. 2002). `core_hawkes_residuals()` returns the rescaled inter-event quantities, which should
then be uniform:

```{r residuals}
U <- core_hawkes_residuals(times, 300, "exponential", fit$theta)
stats::ks.test(U, "punif")$p.value
```

```{r qq}
plot(stats::qunif(stats::ppoints(length(U))), sort(U), pch = 20, cex = .6,
     xlab = "uniform quantiles", ylab = "rescaled residuals",
     main = "Time-rescaling check")
abline(0, 1, col = 2)
```

A misspecified kernel shows as a departure from the diagonal; a p-value well below 0.05
says the model does not explain the clustering it was fitted to.

## Events recorded to a resolution

Administrative data record dates, not instants. Many events on one day are *ties*, and a
Hawkes likelihood with tied times is ill-defined (the kernel at lag zero). `core_hawkes_jitter()`
spreads each day's events uniformly across that day, deterministically from a seed, so the
fit is reproducible and no two events coincide:

```{r jitter}
days <- c(0, 0, 0, 1, 3, 3)
core_hawkes_jitter(days)
anyDuplicated(core_hawkes_jitter(days))
```

Fit on jittered times, and treat the day as the unit of the horizon.

## Evaluating a likelihood directly

`core_hawkes_nll()` returns the negative log-likelihood at given parameters, for profile
plots, likelihood-ratio comparisons or a check against another implementation:

```{r nll}
core_hawkes_nll(times, 300, "exponential", fit$theta)
core_hawkes_nll(times, 300, "exponential", fit$theta * 1.5)
```

## Boundaries and identifiability

A heavy-tailed kernel that tends to its exponential limit has a shape parameter that is not
identified; the fit then stops at the parameter bound and says so (`fit$at_bound`), instead
of reporting an absurd estimate. Short series carry little information about the kernel's
shape: with a few hundred events, compare kernels by AIC but do not read much into the
third decimal of a shape parameter.

## References

Hawkes (1971). Spectra of some self-exciting and mutually exciting point processes.
*Biometrika* 58(1).
Ogata (1981). On Lewis' simulation method for point processes. *IEEE Trans. Inf. Theory*.
Brown, Barbieri, Ventura, Kass, Frank (2002). The time-rescaling theorem and its
application to neural spike train data analysis. *Neural Computation* 14(2).
