---
title: "Key encapsulation: ML-KEM and HQC"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Key encapsulation: ML-KEM and HQC}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(rmoriebricklayer)
```

A key-encapsulation mechanism (KEM) lets two parties agree on a shared secret over a
public channel: the sender *encapsulates* a fresh secret under the recipient's public key,
producing a ciphertext; the recipient *decapsulates* the ciphertext with the secret key and
recovers the same secret. The package implements **ML-KEM** (FIPS 203, lattice-based) and
**HQC** (code-based), natively.

## ML-KEM

```{r mlkem}
key <- kem_keygen(768)
pub <- kem_public_key(key)

# the sender holds only the public key
sent <- kem_encapsulate(pub)
# the recipient recovers the same secret from the ciphertext
got <- kem_decapsulate(key, sent$ciphertext)
identical(sent$shared, got)
```

Three parameter sets trade size for security margin:

```{r sizes}
kem_sizes(512)
kem_sizes(768)
kem_sizes(1024)
```

### Implicit rejection

ML-KEM is designed so that a corrupted ciphertext does **not** produce an error. It produces
a *different*, pseudorandom secret derived from the secret key and the ciphertext. This is
the Fujisaki–Okamoto transform's *implicit rejection*: an attacker who tampers with
ciphertexts learns nothing from the recipient's behaviour, because the recipient behaves
identically either way.

```{r reject}
bad <- sent$ciphertext
substring(bad, 1L, 2L) <- "00"
other <- kem_decapsulate(key, bad)
nchar(other) == 64L          # a 32-byte secret, as always
identical(other, got)        # but not the one the sender has
```

### Masked decapsulation (the default)

Decapsulation is the operation that uses the secret key, so it is the one a power or
electromagnetic side channel would target. Since 0.5.9 it runs **first-order masked** by
default: every value that depends on the secret key is held as two random shares that are
refreshed from the operating system's randomness on every call, and the re-encryption
comparison of the FO transform is done on shares (the decompressed comparison of Bhasin et
al., TCHES 2021). The result is byte-identical to the plain computation:

```{r masked}
identical(kem_decapsulate(key, sent$ciphertext, masked = FALSE), got)
identical(kem_decapsulate(key, bad, masked = FALSE), other)
```

`masked = FALSE` is about five times faster and is the right choice where no physical
attacker is in the threat model (a server in a locked room). The *Side-channel assurance*
vignette explains what the masking defends against and how it was verified.

### Key validation

FIPS 203 §7.3 requires a decapsulation key to be checked before use: its stored hash of the
encapsulation key must match the key it contains. A key that fails this check is refused
rather than used. This check was added after the Wycheproof vectors showed the earlier
code accepted such a key.

## HQC

HQC rests on the hardness of decoding random quasi-cyclic codes, an assumption unrelated to
lattices, which is why NIST selected it as the second KEM. The interface is the same:

```{r hqc}
hkey <- hqc_keygen(1)
hpub <- hqc_public_key(hkey)

hsent <- hqc_encapsulate(hpub)
identical(hqc_decapsulate(hkey, hsent$ciphertext), hsent$shared)
hqc_sizes(1)
```

Two revisions are implemented: the 2025 specification (`version = "v5"`, the default) and
the round-4 submission (`version = "round4"`), which differ in their shared-secret
derivation. A key made under one does not interoperate with the other, and the version is
recorded in the key:

```{r hqc-r4}
old <- hqc_keygen(1, version = "round4")
cap <- hqc_encapsulate(hqc_public_key(old))
identical(hqc_decapsulate(old, cap$ciphertext), cap$shared)
```

### Reproducible keys for tests

Both KEMs accept a 32-byte seed so a test can regenerate the same key. Never use this
outside tests: a key derived from a guessable seed is no key at all.

```{r seed}
identical(hqc_keygen(1, seed = as.raw(1:32))$public,
          hqc_keygen(1, seed = as.raw(1:32))$public)
```

## Using the shared secret

The shared secret is 32 bytes of hex. It is a *key*, not a password: use it directly as the
key of `core_hmac_sha256()` or to derive further keys, never as text to be stored.

```{r use}
tag <- core_hmac_sha256(sent$shared, "the capsule manifest")
core_digest_equal(tag, core_hmac_sha256(got, "the capsule manifest"))
```

## How these implementations are checked

ML-KEM runs against the NIST ACVP and C2SP Wycheproof sets (key generation, encapsulation,
decapsulation and key validation), is compared byte for byte with OpenSSL 3.5, and is
differentially fuzzed against it. Decapsulation's timing is measured on real x86-64 and
arm64 CPUs with dudect, its constant-time property is proved under valgrind, and its masked
arithmetic is assessed for first-order power leakage in a Cortex-M4 simulation. See
`SECURITY.md` in the repository.
