---
title: "Missingness, duplicates, outliers and distinct counts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Missingness, duplicates, outliers and distinct counts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(rmoriebricklayer)
```

Before a published table is analysed it should be looked at for the four things that most
often go wrong in ingestion: missing values with a pattern, duplicated rows, rows that are
ordinary on each column but impossible jointly, and identifier columns with fewer distinct
values than they should have.

## Missingness

```{r missing}
df <- data.frame(a = c(1, NA, 3), b = c(NA, NA, 3), c = 1:3)
missingness_summary(df)
missingness_summary(data.frame(x = 1:3, y = 4:6))
```

A pattern matters more than a rate: `missingness_pattern()` tabulates the combinations of
missing columns, `missing_runs()` finds runs of consecutive missing rows (a feed that
stopped), and `missingness_map()` draws an inline text map of where the gaps are.

```{r pattern}
set.seed(1)
big <- data.frame(id = 1:40, v = stats::rnorm(40), w = stats::rnorm(40))
big$v[c(10:14, 30)] <- NA
big$w[c(10:14)] <- NA
missingness_pattern(big)
missing_runs(big)
```

Five rows missing both `v` and `w` in a block is a different problem from six scattered
gaps; the first points at the source, the second at the records.

## Duplicates

```{r dup}
d <- data.frame(id = c(1, 2, 2, 3, 3, 3), x = c("a", "b", "b", "c", "c", "d"))
duplicate_rows(d)
duplicate_rows(d, columns = "id")
```

Full-row duplicates are usually an ingestion artefact; duplicates on the key alone are
usually a data-model finding (the key is not a key).

## Multivariate outliers

A row can be inside every column's range and still be impossible: short and heavy, young
and retired. The Mahalanobis distance measures how far a row is from the centre of the
joint distribution in the metric of its covariance. The **robust** version uses the median
and the MAD, so a cluster of bad rows cannot drag the centre toward itself and hide:

```{r maha}
set.seed(1)
n <- 200
df <- data.frame(height = stats::rnorm(n, 170, 10))
df$weight <- df$height * 0.5 + stats::rnorm(n, 0, 5)
df[1, ] <- list(height = 155, weight = 100)      # ordinary on each, unlikely jointly

out <- mahalanobis_outliers(df)
head(out, 3)
out$row[1] == 1
range(df$height[-1])
range(df$weight[-1])
```

With a cluster of bad rows, the classical distance of a bad row is *smaller* than the robust
one, because the cluster has pulled the classical centre toward itself:

```{r maha-robust}
bad <- df
bad[2:21, ] <- list(height = 150, weight = 110)
d_classical <- mahalanobis_outliers(bad, robust = FALSE)
d_robust <- mahalanobis_outliers(bad, robust = TRUE)
d_classical$distance[d_classical$row == 2] < d_robust$distance[d_robust$row == 2]
```

The flag threshold is a chi-square quantile (`alpha = 0.001` by default) with the number of
columns as degrees of freedom. A duplicated column has no distance defined and is refused.

## Distinct counts in fixed memory

How many distinct identifiers does a 200-million-row file have? Holding them all in a set
needs memory proportional to the answer. A HyperLogLog sketch answers to within about 1% in
a few kilobytes, can be fed block by block, and two sketches of two files merge into one:

```{r hll}
set.seed(1)
x <- sample(1:5000, 200000, replace = TRUE)
distinct_count(distinct_sketch(x))
length(unique(x))

s <- distinct_sketch(x[1:100000])
s <- distinct_sketch(x[100001:200000], registers = s)
distinct_count(s)

a <- distinct_sketch(x[1:100000])
b <- distinct_sketch(x[100001:200000])
distinct_count(sketch_merge(a, b))
```

Small cardinalities are near-exact through linear counting, and an empty input has no
distinct values:

```{r hll-small}
distinct_count(distinct_sketch(c("a", "b", "c", "a", "b")))
distinct_count(distinct_sketch(character(0)))
```

An identifier column whose distinct count is far below its row count is not an identifier;
`capsule_drift()` uses the same reasoning to decide which columns to compare as categories.

## Does the data look fabricated or truncated?

Benford's law gives the expected distribution of leading digits for a quantity that spans
several orders of magnitude; a strong departure is a reason to ask how the numbers were
produced (rounding, capping, invention), never a proof of anything on its own:

```{r benford}
set.seed(3)
benford_test(10^stats::runif(2000, 0, 6))$p_value
benford_test(as.numeric(paste0(sample(1:9, 2000, TRUE), "000")))$p_value
```

## Order of operations

1. `missingness_summary()` and `missingness_pattern()`; decide whether gaps are structural.
2. `duplicate_rows()` on the full row and on the key.
3. `mahalanobis_outliers()` on the numeric columns, robust.
4. `distinct_sketch()` on identifiers when the table is large.
5. Then, and only then, the analysis.
