---
title: "Post-quantum signatures: ML-DSA, SLH-DSA and XMSS"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Post-quantum signatures: ML-DSA, SLH-DSA and XMSS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(rmoriebricklayer)
```

A capsule is only as trustworthy as the signature on its manifest. `rmoriebricklayer`
implements the three NIST post-quantum signature standards natively, with no system
cryptography library: **ML-DSA** (FIPS 204, lattice-based), **SLH-DSA** (FIPS 205,
hash-based, stateless) and **XMSS** (RFC 8391, hash-based, stateful). All three are used
through the same two calls, `capsule_sign()` and `capsule_verify()`.

## Why post-quantum, and why three schemes

A signature made today may need to verify in twenty years, when a large quantum computer
could break RSA and ECDSA with Shor's algorithm. The three schemes rest on different
assumptions, so a break of one does not take the others:

* **ML-DSA** (Module-Lattice Digital Signature Algorithm, formerly Dilithium) rests on the
  hardness of Module-LWE and Module-SIS. Small keys and signatures, fast. The default.
* **SLH-DSA** (Stateless Hash-based, formerly SPHINCS+) rests only on the security of a hash
  function. Conservative; its signatures are large.
* **XMSS** is also hash-based but *stateful*: every key signs a fixed number of times and
  an index must never be reused. The package carries the state forward for you.

## ML-DSA: the default

```{r mldsa}
key <- fips_keygen("ML-DSA-65")
key$scheme

sig <- capsule_sign("a manifest digest", key)
capsule_verify("a manifest digest", sig, fips_public_key(key))
```

The verifier needs only the public key. Any change to the message is caught:

```{r mldsa-tamper}
capsule_verify("a manifest digest (edited)", sig, fips_public_key(key))
```

### Context strings bind a signature to its purpose

FIPS 204 lets a signature carry a *context*: a short string hashed in with the message. The
same message signed for one purpose does not verify under another, which stops a signature
made for "staging" being replayed as a "release" approval:

```{r context}
sig2 <- capsule_sign("a manifest digest", key, context = "release")
capsule_verify("a manifest digest", sig2, key, context = "release")
capsule_verify("a manifest digest", sig2, key, context = "staging")
```

### Hedged and deterministic signing

By default ML-DSA signing is *hedged*: fresh randomness enters every signature, so two
signatures of one message differ (FIPS 204 recommends this; it also protects against
fault attacks on the nonce). `deterministic = TRUE` gives the variant where the signature
depends only on the key and the message:

```{r deterministic}
s1 <- capsule_sign("same message", key, deterministic = TRUE)
s2 <- capsule_sign("same message", key, deterministic = TRUE)
identical(s1$signature, s2$signature)

h1 <- capsule_sign("same message", key)
h2 <- capsule_sign("same message", key)
identical(h1$signature, h2$signature)
```

### Signing a digest computed elsewhere

ML-DSA signs `mu = H(tr || M')`, a digest of the message that depends on the public key but
not on the secret. `fips_mu()` computes it from the *public* key, so the machine holding the
secret never needs to see the message:

```{r mu}
mu <- fips_mu(fips_public_key(key), "a manifest digest", context = "release")
sig3 <- fips_sign_mu(key, mu)
fips_verify_mu(key, mu, sig3)
# the same signature verifies the ordinary way
capsule_verify("a manifest digest", sig3, key, context = "release")
```

Since 0.5.9, ML-DSA signing runs **first-order masked** by default (see the
*Side-channel assurance* vignette): every secret-dependent value is held as two random
shares. The signatures are byte-identical to the unmasked computation; only the cost differs.

## SLH-DSA: hash-based, stateless

```{r slh}
fips_sizes("ML-DSA-65")
fips_sizes("SLH-DSA-SHAKE-128f")
```

The price of dropping the lattice assumption is size: an SLH-DSA signature is far larger
than an ML-DSA signature at the same security level.

```{r slh-size}
fips_sizes("SLH-DSA-SHAKE-128f")[["signature"]] > fips_sizes("ML-DSA-44")[["signature"]]
```

The calls are identical:

```{r slh-sign}
hkey <- fips_keygen("SLH-DSA-SHAKE-128f")
hsig <- capsule_sign("a manifest digest", hkey)
capsule_verify("a manifest digest", hsig, fips_public_key(hkey))
```

### Pre-hash variants

HashML-DSA and HashSLH-DSA sign a digest of the message instead of the message itself,
for very large inputs or when the message is hashed on another machine. The hash
function's OID is bound into the signature, so two digests of the same length never
verify for each other:

```{r prehash}
p <- capsule_sign("a large artifact", key, prehash = "sha512")
capsule_verify("a large artifact", p, key, prehash = "sha512")
capsule_verify("a large artifact", p, key, prehash = "sha3_512")
```

## XMSS: hash-based, stateful

An XMSS key of height *h* signs 2^h messages. The package stores the next index in the key
and returns the advanced key with every signature; reusing an index would break security,
so always carry `key_state` forward:

```{r xmss}
xkey <- pqc_keygen(height = 3)
xkey$capacity
xkey$next_index

s1 <- capsule_sign("manifest-1", xkey)
capsule_verify("manifest-1", s1, signing_public_key(xkey))

xkey <- s1$key_state
xkey$next_index
s2 <- capsule_sign("manifest-2", xkey)
capsule_verify("manifest-2", s2, signing_public_key(xkey))

# a signature does not transfer to another message
capsule_verify("manifest-1", s2, signing_public_key(xkey))
```

## Every way of being wrong returns FALSE

Verification never errors on malformed input; it returns `FALSE`. A truncated or edited
signature, a foreign key and an edited message all fail the same way:

```{r wrong}
pub <- fips_public_key(key)
bad <- sig
bad$signature <- paste0("ff", substring(bad$signature, 3))
capsule_verify("a manifest digest", bad, pub)
trunc <- sig
trunc$signature <- substring(sig$signature, 1, 64)
capsule_verify("a manifest digest", trunc, pub)
capsule_verify("a manifest digest", sig, fips_public_key(fips_keygen("ML-DSA-65")))
```

## How these implementations are checked

Every scheme is compared byte for byte with OpenSSL 3.5 and run against the NIST ACVP and
C2SP Wycheproof vector sets on every change (6,848 checks, 0 failures at 0.5.9), fuzzed
differentially against OpenSSL, and run under valgrind with the secret key marked undefined
to prove no branch or memory address depends on it. `SECURITY.md` in the package
repository has the full table of what each check catches and what it does not.
