---
title: "Side-channel assurance: constant time, timing tests, masking and simulated power analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Side-channel assurance: constant time, timing tests, masking and simulated power analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(rmoriebricklayer)
```

Correct cryptography can still leak its key through *how* it computes: how long an
operation takes, how much power the chip draws while it runs, what it radiates. This
vignette explains what the package does about each channel, how that was verified, and
exactly what remains outside the verification. The harnesses live under `inst/` in the
installed package and run in continuous integration on every change.

```{r harnesses}
list.files(system.file(package = "rmoriebricklayer"), pattern = "^(ctcheck|dudect|tvla|fuzz)$")
```

## 1. Timing

### The threat
An operation's duration depends on a secret when a branch condition depends on it, when a
memory address depends on it (cache timing), or when an instruction's latency depends on
its operand (division on many cores). An attacker who can time the operation, locally or
across a network, recovers the secret bit by bit.

### Static proof: ctgrind (`inst/ctcheck`)
The secret key is marked *undefined* to valgrind memcheck (Langley's ctgrind, 2010).
Memcheck then reports every conditional jump and every memory address that depends on
undefined data, which is exactly the set a timing attacker can observe. A clean run is a
proof for that binary, for every input of that length. The only branches allowed are the
ones the standards publish by design (a rejected sample in ML-DSA, the challenge, the
hints), each passed through an explicit declassification point. The harness is built from
the package's own sources with GCC and Clang, and it also scans the dead stack after each
operation for surviving copies of the secret.

```sh
bash inst/ctcheck/build.sh     # needs R, a C++17 compiler, valgrind headers
bash inst/ctcheck/run.sh       # exit 0 = every case clean
```

### Measured on hardware: dudect (`inst/dudect`)
Reparaz, Balasch and Verbauwhede's *Dude, is my code constant time?* (DATE 2017): each
measurement times the operation on an input drawn at random from a *fixed* class or a
*random* class; Welch's t compares the two timing distributions. |t| > 10 is a real
difference; |t| < 4.5 is none. A positive control, an early-exit comparison, must be flagged
or the run is invalid. CI runs it on x86-64 Linux, arm64 Linux and Apple silicon.

```sh
bash inst/dudect/build.sh
N=300000 bash inst/dudect/run.sh     # exit 1 on any |t| > 10
```

### What this found
ML-KEM compression divided by *q*; at `-O0` the compiler emitted a hardware divide whose
latency depends on the operand (the KyberSlash class of bug). It is a multiply-and-shift
now. Three "add *q* when negative" corrections had compiled to conditional execution on a
secret share's sign; they are branch-free masks now.

## 2. Power and electromagnetic emanation

### The threat
A CMOS circuit draws power when its transistors switch. Under the **Hamming-weight model**
a power sample is proportional to the number of 1-bits in a value being written; under the
**Hamming-distance model** it is proportional to the number of bits that *flip* in a
register or bus between its old and new value. Either lets an attacker correlate recorded
traces against hypotheses on a secret-dependent intermediate (differential and correlation
power analysis). Constant-time code is fully exposed to this: the values are secret even
when the timing is not. EM attacks read the same switching through a near-field probe.

### The countermeasure: first-order masking
Every secret-dependent intermediate *x* is split into two shares that are each uniformly
random on their own, `x = x0 XOR x1` (Boolean) or `x = x0 + x1 mod q` (arithmetic). Linear
operations act share by share; non-linear ones consume fresh randomness so that no single
intermediate depends on the secret. In the probing model of Ishai, Sahai and Wagner (2003)
this is *first-order security*: any one observed value is independent of the secret. The
cost an attacker then pays grows exponentially in the masking order with the measurement
noise as the base.

ML-KEM decapsulation and ML-DSA signing are masked by default. The output is byte-identical
to the unmasked computation, which the full vector suite checks on every vector:

```{r agree}
key <- kem_keygen(512)
sent <- kem_encapsulate(kem_public_key(key))
identical(kem_decapsulate(key, sent$ciphertext, masked = TRUE),
          kem_decapsulate(key, sent$ciphertext, masked = FALSE))

skey <- fips_keygen("ML-DSA-44")
sig <- capsule_sign("manifest", skey, deterministic = TRUE)
capsule_verify("manifest", sig, fips_public_key(skey))
```

The gadgets (`src/rmbl_masked.h`) are the published ones: Trichina's two-share AND; a
bitsliced ripple-carry adder over Boolean shares (Goubin 2001; Bronchain and Cassiers
2022); arithmetic-to-Boolean conversion by the decoding of Bos et al. (TCHES 2021) and the
decompressed ciphertext comparison of Bhasin et al. (TCHES 2021); Boolean-to-arithmetic
conversion by masked modular subtraction; masked centred-binomial sampling; and the masked
Keccak of Bertoni et al. (2010). For ML-DSA, `HighBits(w)` is public (it is hashed into the
challenge) and is revealed one bit at a time; the three rejection tests reveal one bit each.

### Why the share-pair operations are assembly
A C compiler may place share 0 and then share 1 of one value in the same register. Under
the Hamming-distance model that register's flips are `HW(x0 XOR x1) = HW(x)`: the secret, in
the clear. The simulator below showed exactly this for the C gadgets (|t| up to 135) while
the Hamming-weight model reported them clean. So every operation that touches both shares
is inline assembly (Thumb-2, x86-64, aarch64) with a fixed register order, every temporary
and destination zeroed before use, a zero stored between the share-0 and share-1 stores,
and the registers scrubbed between passes that handle one share at a time. On any other CPU
the same operations run as plain C, with the same results and without the register
discipline.

### Simulated power analysis: TVLA (`inst/tvla`)
The target is the Cortex-M4, the reference core of the embedded post-quantum literature.
The package's own headers are compiled for it and executed instruction by instruction in
the unicorn emulator; capstone decodes which registers each instruction writes, and one
sample per instruction is recorded under both leakage models. The statistic is TVLA
(Goodwill et al. 2011; ISO/IEC 17825): a fixed secret against random secrets, Welch's t at
every sample, in two independent experiments; a point leaks when |t| > 4.5 in both with the
same sign. Each masked kernel runs beside its unmasked control (the same code with share 1
zero), because a control that does not leak means the harness sees nothing.

```sh
bash inst/tvla/build.sh                 # arm-none-eabi-g++, unicorn, capstone
cd inst/tvla && ./build/tvla all 5000   # Hamming-weight model
TVLA_MODEL=hd ./build/tvla all 5000     # Hamming-distance model
```

At 2 x 3,000 traces every masked kernel is clean under both models (max |t| 3.0) and every
unmasked control leaks (|t| 87 to 163). CI fails on a masked leak under either model. The
full table is in `inst/tvla/README.md`.

```{r tvla-readme}
readLines(system.file("tvla", "README.md", package = "rmoriebricklayer"), n = 12)
```

## 3. What is outside these checks

Honesty about scope is part of the assurance. The checks do **not** model:

* **Electromagnetic emanation** on a specific board. The simulation models switching
  activity, which is the source of EM leakage too, but a measurement needs a probe.
* **Fault injection**: glitching the clock or voltage to skip an instruction. The checks
  assume the program runs as written.
* **Glitches and coupling** a leakage model omits: pipeline forwarding, write-back buffers
  and capacitive coupling between wires, which can combine shares (the MIRACLE study,
  TCHES 2021, catalogues them).
* **Higher-order attacks**: two shares resist one probe; a second-order attacker combines
  two leakage points.

Each has a software route (richer fitted leakage models such as ELMO, a fault campaign in
the same emulator, higher-order gadgets, formal verification of the gadgets with maskVerif)
and, for EM and for confirming a given chip, a hardware route. The repository's
`SECURITY.md` states the current boundary; the release bundle includes ChipWhisperer
firmware and a capture script for anyone with a board.

## 4. Choosing the plain path

Masking costs about 5x on decapsulation and 16x on signing (13 ms for ML-DSA-65 on a
laptop). Where the threat model has no physical attacker, `kem_decapsulate(masked = FALSE)`
keeps the plain path. Timing protections cannot be switched off; they are not optional.
