---
title: "Parsing SIU director's reports"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Parsing SIU director's reports}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(rmoriebricklayer)
```

Ontario's Special Investigations Unit publishes a director's report for every investigation
as an HTML page. The reports share a structure but not a template: fields appear under
varying headings, counts are written in words or figures, and the text is long. The package
parses them into a fixed, panel-reviewed schema with a parser written for hostile input.

## The schema

```{r schema}
sch <- bricklayer_siu_schema()
head(sch$name, 20)
nrow(sch)
```

## Parsing a report

The package ships a synthetic report with the structure of a real one and no real person
in it:

```{r parse}
f <- bricklayer_parse_siu(system.file("extdata", "siu_synthetic_report.html",
                                      package = "rmoriebricklayer"))
f[["number_of_subject_officials"]]
f[["date_of_incident_iso"]]
```

## The pieces

HTML to plain text, with the layout tags turned into line structure and entities decoded:

```{r text}
bricklayer_siu_text("<p>Number of SIU Investigators assigned: 3</p>")
```

The subject-official count, resolved from the enumerated list rather than a headline number
that reports sometimes omit:

```{r so}
bricklayer_siu_resolve_so("Subject Officials\nSO #1 Interviewed\nSO #2 Declined interview")
```

Dates in the report's prose, normalised to ISO:

```{r dates}
bricklayer_siu_iso_date("March 4, 2024")
bricklayer_siu_iso_date("not a date")
```

## Written for hostile input

A parser that reads 5,000 public web pages will meet broken HTML, megabyte-long lines,
unclosed tags and entities of every shape. Three review rounds on this parser established
and tested the following, each pinned by a test over the sources:

* **No regex crosses a line without a bound.** Two passes that once scanned to the end of the
  document on `[\s\S]*?` were replaced by `find()` scans after 44 KB of period-free
  boilerplate overflowed the C stack; a hygiene test now asserts no unbounded cross-line
  regex exists in the SIU sources.
* **Every pass is linear.** Two tag-stripping passes were quadratic (re-scanning from each
  unclosed `<script` or bare `<`); at the 2 MiB input cap one took 40 minutes. Both are
  single-pass now.
* **Lines are capped at 2,000 characters** and split at the last sentence end inside the
  window, else the last space; the functions warn how many lines were split, because a field
  spanning a split can come back incomplete.
* **Numbers are bounded.** Every string-to-integer conversion goes through one helper that
  reads up to nine digits; `SO #1234567` is 1234567, and `SO #4444444444444444` is ignored
  as a tag (the count falls back to the other cues),
  not an overflow.
* **The user can interrupt.** The whole path polls for an interrupt through a hook the host
  installs, and raises it only from a frame that owns no live C++ objects.
* **Input is capped at 2 MiB**, and the fuzzers run the parser under an 8 MB stack like R's,
  with seeds for every shape the reviewers used.

## Fetching live reports

`bricklayer_fetch_siu()` fetches a report by its identifier through the package's checked
transport (address validation, redirect policy, resource limits) with an Internet Archive
fallback, and `bricklayer_fetch_parse_siu()` fetches and parses in one call:

```{r fetch, eval = FALSE}
f <- bricklayer_fetch_parse_siu("24-OCI-123")
```

The corpus of parsed reports, reviewed round by round, is distributed as a dataset by the
companion package `rmoriedata`; this package provides the parser so the corpus can be
rebuilt and audited.
