---
title: "Stock and flow: population, admissions and length of stay"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Stock and flow: population, admissions and length of stay}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(rmoriebricklayer)
```

A custodial or clinical population can be described two ways: the **stock** (how many are
held on an average day) and the **flow** (how many pass through in a period). The two are
linked by the length of stay, and a published table that reports one without the others is
easy to misread. These functions implement the standard measures (Lakner's formulation, as
used in the Canadian corrections statistics) with exact intervals.

## Average daily population

The stock. Detention days in a period divided by the days in the period:

```{r adp}
# Lakner's worked example: 3,000 inmates served 13,500 detention days in a year
adp(13500)

# per-person days over a 30-day period give the same total
adp(c(10, 20, 30), t = 30)
```

`adp_from_counts()` estimates person-days from periodic headcounts (a monthly census) when
daily data are not published.

## Average length of stay

Detention days divided by the number of people who completed a stay:

```{r alos}
# Lakner: 2,700 inmates admitted and released served 12,150 detention days
alos(12150, 2700)
```

## Admissions implied by stock and stay

Little's law ties the three together: in a stationary system, stock = flow × stay, so
admissions = ADP × days / ALOS. Lakner (p. 20) runs it the other way: an average daily
population of 25 and 1,750 admissions in a year imply a stay of 5.2 days, and back again:

```{r admissions}
alos_implied <- 25 * 365 / 1750
round(alos_implied, 1)
admissions(25, alos_implied)
```

## Stock and flow together

`stock_flow()` takes detention days and people per period, with an exposure (the population
at risk) to turn counts into rates, and reports how the stock, the flow and the stay moved
between periods. The pattern it is built to surface: **fewer people, held longer**, so the
flow falls while the stock rises.

```{r stockflow}
stock_flow(
  days = c(115674, 126121), people = c(12647, 9608),
  period = c("2023", "2025"),
  exposure = c(15495050, 16256538)
)
```

A table that reported only admissions would show a decline; one that reported only the
daily count would show growth. Both are true, and the length of stay is the reason.

## The distribution of stays

The mean stay is pulled far above the median by a few long stays, so the summary that
matters is the whole distribution:

```{r stays}
stays <- c(rep(1, 40), rep(3, 30), rep(10, 20), 60, 90, 120)
stay_summary(stays)
stay_summary(stays)[, c("mean", "median", "max")]
```

`stay_summary()` gives the mean with its interval, the quantiles and the maximum, so a
report can say "half of stays were three days or less, and the longest was 120" instead of
"the average stay was 9.6 days", which describes almost no one.

## Periods and fiscal years

Published series use fiscal years and uneven periods:

```{r periods}
period_days("2024-04-01", "2025-03-31")
fiscal_year_label(2024)
```

## From a published table

1. `adp()` for the stock from detention days, or `adp_from_counts()` from headcounts.
2. `alos()` for the stay; `admissions()` to check the implied flow against the published one.
3. `stock_flow()` across periods, with the population as exposure.
4. `stay_summary()` whenever per-person stays are available, and report the median.

## Reference

Lakner, *A Manual of Correctional Statistics* (the ADP and ALOS worked examples at pp. 15
and 17 are reproduced above).
