---
title: "Statistics for a published administrative table"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Statistics for a published administrative table}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(rmoriebricklayer)
```

An open-data extract from a criminal-justice system has a particular
shape: a handful of fiscal years, counts rather than measurements,
categories reported as bands rather than values, and a region code with
no geometry attached. The general-purpose toolkits assume none of that.
This vignette covers the functions that do.

The running example is the shape Ontario's inmate datasets have -- one
row per placement, keyed on the fiscal year's end year, with banded
counts and a region.

## Bands

Published categories are intervals. Anything computed from them rests on
an assumption about where inside each band the mass sits, and on a
second assumption about where the open top band ends.

```{r}
bands <- parse_bands(c("1", "2 to 5", "6 to 10", "Greater than 10"))
bands
```

Note what the bounds say. `"Greater than 10"` starts at 11, because the
quantity is a count of placements; `"10 or more"` would start at 10. One
whole placement separates the two readings, and only the label says
which was meant.

The representative value needs a rule, and the open band needs a cap:

```{r}
band_values(bands)
```

The last row is marked `assumed`, because a band with no upper bound has
no midpoint. Everything downstream inherits that assumption, so measure
how much it matters:

```{r}
counts <- c(1200, 430, 110, 38)
band_sensitivity(bands, counts)
```

Read the span, not the middle row. A statistic that moves by more than a
tenth of itself across plausible caps is a property of the assumption as
much as of the data, and the honest report is the range.

## Concentration

The recurring question is whether a few individuals account for most of
the total.

```{r}
placements <- expand_bands(bands, counts, open_upper_cap = 25)
gini(placements)
top_share(placements, c(0.01, 0.05, 0.1))
```

Gini has a ceiling that depends on the number of units: for `n` units the
maximum is `1 - 1/n`, not 1. A Gini of 0.9 means something different
across ten units than across ten thousand.

```{r}
c(ten = 1 - 1 / 10, thousand = 1 - 1 / 1000)
```

If the tail looks heavy, fit it rather than asserting it -- and read the
goodness-of-fit distance before the exponent:

```{r}
fit <- hill_tail_index(placements, x_min = 1)
c(alpha = round(fit$alpha, 2), ks = round(fit$ks, 3),
  reliable = fit$reliable)
```

A large `ks` means the tail is not a power law, whatever `alpha` came out
as. The estimator maximises the exact discrete likelihood: the
closed-form continuity correction usually quoted is an asymptotic
approximation in the threshold, and at a threshold of one -- where counts
start -- it is badly biased.

## Trend across a handful of years

Five annual points support a rank test and a resistant slope. They do
not support a model with an autocorrelation structure.

```{r}
y <- c(402, 377, 190, 268, 331)
tt <- trend_test(y)
c(tau = tt$tau, p = round(tt$p_value, 4), slope = tt$slope)
tt$method
```

The p-value is exact, by enumerating all 120 orderings. For counts, the
trend is better expressed as a multiplicative change per year:

```{r}
ct <- count_trend(y)
c(rate_ratio = round(ct$rate_ratio, 3),
  lower = round(ct$lower, 3), upper = round(ct$upper, 3),
  dispersion = round(ct$dispersion, 2))
```

The dispersion is reported because a Poisson interval assumes it is one.
Above 1.5 the interval is widened to the quasi-Poisson one rather than
being left too narrow.

### Did it change when the policy changed?

```{r}
sc <- step_change(c(100, 104, 98, 60, 63, 58), x = 2018:2023)
c(after = sc$break_after, before = sc$before, after_mean = sc$after,
  p = round(sc$p_value, 4))
```

The p-value comes from the permutation distribution of the **maximum**
over splits, not from the best split's own test. That distinction is the
difference between a change-point test and a way of finding a break in
noise. It also has a consequence worth stating plainly: six points
cannot produce a p-value below about 0.1, however clean the step, because
only 72 of the 720 orderings separate three low values from three high
ones.

```{r}
(1 + 2 * factorial(3)^2) / (1 + factorial(6))
```

## Regions

A region code is an areal unit, not a coordinate. Three things go wrong
with a table of regional counts: the regions hold different numbers of
people, they hold different *kinds* of people, and a small region's rate
is mostly noise.

```{r}
d <- expand.grid(
  region = c("Central", "Eastern", "Northern", "Toronto", "Western"),
  age = c("18 to 24", "25 to 49", "50+"),
  stringsAsFactors = FALSE
)
d$pop <- rep(c(4000, 3000, 800, 5000, 2500), 3) *
  rep(c(0.3, 0.55, 0.15), each = 5)
set.seed(13)
rate <- c(Central = 0.006, Eastern = 0.011, Northern = 0.010,
          Toronto = 0.016, Western = 0.012)
d$n <- rpois(nrow(d), lambda = d$pop * rate[d$region])
```

Indirect standardisation handles the second problem: the expected count
is what each region would have under the overall rate in every age band,
given its own age composition.

```{r}
e <- expected_counts(d$n, d$pop, d$region, strata = d$age)
e
```

The expected counts total the observed ones. That identity is what makes
them a standardisation rather than a prediction.

```{r}
c(observed = sum(e$observed), expected = sum(e$expected))
```

The ratio, with an exact interval -- the counts here are small, and a
normal approximation on a count of six is a decoration rather than an
interval:

```{r}
sir(e$observed, e$expected, e$area)
```

And the third problem. A league table of rates ranks the small regions to
both ends by construction, so borrow strength across regions:

```{r}
eb_rates(e$observed, e$expected, e$area)
```

`shrinkage` is how far each estimate was pulled toward the overall rate.
It is governed by the expected count, which is to say by how much
information the region carries -- not by how extreme its ratio was.

The funnel plot shows the reader the same thing directly: what range a
region's ratio could take, given its size, if it were no different from
anywhere else.

```{r}
funnel_limits(c(5, 20, 50, 200))
```

### Autocorrelation, if you have a neighbour list

```{r}
nb <- list(c(2L, 4L), c(1L, 3L), c(2L, 5L), c(1L, 5L), c(3L, 4L))
set.seed(1)
mi <- morans_i(e$observed / e$expected, nb, n_perm = 999L)
c(I = round(mi$I, 3), expectation = round(mi$expectation, 3),
  p = mi$p_value)
```

The neighbour list has to be supplied. An extract keyed on a region ships
no geometry, and a guessed adjacency would make the answer a property of
the guess. Note too that the null expectation is `-1/(n - 1)`, not zero:
with five regions, a small negative `I` is what independence looks like.

```{r}
-1 / (5 - 1)
```
