---
title: "Trends and change in short count series"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Trends and change in short count series}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(rmoriebricklayer)
```

A published series is usually short: five to ten annual counts. Three questions recur. Is
there a monotone trend? How fast is the count changing per period? Did something change at
one point? Each has a method suited to short series, and each is wrong to answer with a
least-squares line on five points.

## A monotone trend: Mann–Kendall and Sen's slope

The Mann–Kendall test counts, over all pairs of periods, how many later values exceed
earlier ones minus how many fall below. It is distribution-free and exact for small *n*,
so it is informative where a regression's standard error would be nearly meaningless.
Sen's slope is the median of all pairwise slopes, a robust rate of change.

```{r mk}
y <- c(402, 377, 190, 268, 331)        # five years of placements
trend_test(y)
```

A monotone series is detected even at *n* = 5:

```{r mk-mono}
trend_test(c(1, 2, 3, 4, 5))
```

One aberrant period does not create a trend:

```{r mk-outlier}
trend_test(c(100, 100, 100, 100, 900))$p_value
```

From a data frame, with a direction when one is hypothesised in advance:

```{r mk-df}
d <- data.frame(year = 2019:2023, n = y)
trend_test(d, value = "n", period = "year")$slope
trend_test(d, value = "n", period = "year", alternative = "decreasing")$p_value
```

## The rate of change: a Poisson trend

For counts, the natural model is a Poisson log-linear trend: `log E[y] = a + b * t`, so
`exp(b)` is the **rate ratio per period**, the multiplicative change from one period to the
next. `count_trend()` fits it with an interval:

```{r poisson}
set.seed(3)
y <- stats::rpois(8, lambda = 200 * 0.85^(0:7))   # falling ~15% a year
fit <- count_trend(y)
round(fit$rate_ratio, 3)
```

With a varying denominator the trend is in the *rate*, via an offset:

```{r offset}
count_trend(c(20, 25, 30), offset = c(1000, 1500, 2500))$rate_ratio
```

Counts here rise while the rate per 1,000 falls, which is the finding.

## A step change

A trend test asks "is it drifting?"; a step-change test asks "did it jump once?". The best
single split point is found by maximising the between-segment difference, and its
significance is judged by permutation, which accounts for having searched over every
possible split:

```{r step}
step_change(c(100, 104, 98, 60, 63, 58))
```

Pure noise still yields a best split, and it is not significant:

```{r step-noise}
set.seed(2)
step_change(stats::rnorm(12))$p_value
```

## Choosing between them

| Question | Tool | Reports |
|---|---|---|
| Is the series moving in one direction? | `trend_test()` | Mann–Kendall p-value, Sen's slope with interval |
| By what factor per period? | `count_trend()` | rate ratio with interval, optional offset |
| Did it change once, and when? | `step_change()` | split point, segment means, permutation p-value |
| Year over year, with publication rounding | `yoy()` (its own vignette) | exact intervals on each change |

All of them work on the five to ten values a published table gives, and none of them
assumes the counts are large.

## References

Mann (1945). Nonparametric tests against trend. *Econometrica* 13(3).
Kendall (1975). *Rank Correlation Methods*. 4th ed.
Sen (1968). Estimates of the regression coefficient based on Kendall's tau. *JASA* 63(324).
