---
title: "Validation rules, inferred schemas and synthetic stand-ins"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Validation rules, inferred schemas and synthetic stand-ins}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(rmoriebricklayer)
```

A capsule says what shape its data must have. This vignette covers the three tools for
stating and checking that shape: explicit **rules**, a **schema** inferred from a known-good
table, and a **synthetic stand-in** generated from the schema when the portal is unreachable.

## Rules

A rule is a name, a predicate and optionally a column. Row-wise rules report how many rows
fail and which; table-wise rules take the whole data frame. A `fatal` severity stops the
pipeline; a `warning` is recorded.

```{r rules}
df <- data.frame(id = c(1, 2, 2), age = c(30, -5, 40),
                 start = c(1, 5, 3), end = c(2, 4, 9))

rules <- list(
  rule("age_non_negative", function(v) v >= 0, column = "age"),
  rule("id_unique", function(v) !anyDuplicated(v), column = "id", severity = "fatal"),
  rule("dates_ordered", function(d) all(d$start <= d$end))
)

issues <- validate_rules(df, rules)
names(issues)
issues$age_non_negative$n_failed
issues$age_non_negative$rows
```

Clean data produces nothing:

```{r rules-clean}
clean <- data.frame(id = 1:3, age = c(30, 31, 40), start = 1:3, end = 4:6)
length(validate_rules(clean, rules))
```

Two behaviours matter for trust. A rule for an absent column is **skipped**, not failed: a
missing column is the schema's finding to report. And a rule that *errors* is a **failure**,
never a silent pass:

```{r rules-edge}
length(validate_rules(data.frame(id = 1:3), rules[1:2]))
broken <- list(rule("bad", function(v) stop("boom"), column = "age"))
validate_rules(clean, broken)$bad$message
```

### Ready-made rules

Common checks have constructors so the intent is in the name:

```{r ready}
rules2 <- list(
  rule_not_null("id"),
  rule_unique("id", severity = "fatal"),
  rule_between("age", 0, 120),
  rule_in_set("grade", c("a", "b", "c")),
  rule_regex("code", "^[A-Z]{2}-[0-9]{4}$"),
  rule_within_n_mads("score", n = 3),
  rule_col_count(6, severity = "fatal")
)
length(rules2)
```

`rule_increasing()`, `rule_distinct_rows()` and `rule_complete_rows()` complete the set.
`apply_schema_validation()` runs a rule list and stops on any fatal issue, which is the call
a pipeline makes.

## Inferring a schema

Given a table known to be right, `infer_schema()` pins what can be pinned: column names and
types, numeric ranges with a little slack, and the level set of categorical columns (up to
`max_levels`, so free text is not mistaken for a code list):

```{r infer}
set.seed(1)
df <- data.frame(
  id = 1:100,
  score = stats::runif(100, 0, 10),
  grade = sample(c("a", "b", "c"), 100, TRUE),
  note = paste0("free text ", 1:100),
  stringsAsFactors = FALSE
)
sch <- infer_schema(df)
sch$expected_value_sets
```

It validates the table it was learned from, and catches a vanished column or a new
category:

```{r infer-check}
length(validate_schema(df, list(schema = sch)))
length(validate_schema(df[, -3], list(schema = sch))) > 0
bad <- df
bad$grade[1] <- "z"
length(validate_schema(bad, list(schema = sch))) > 0
```

A schema goes into the capsule's provenance record; `validate_schema()` is what
`verify_capsule()` runs on a fresh download.

## Synthetic stand-ins

When the portal is down, the pipeline should still run end to end on data of the right
shape, clearly marked as synthetic. A recipe names each column's generator:

```{r synthetic}
recipe <- list(
  n_rows = 20, seed = 42,
  columns = list(
    year   = list(type = "sample", values = list(2024, 2025)),
    alert  = list(type = "bernoulli", p = 0.2),
    visits = list(type = "poisson", lambda = 3, min = 1),
    id     = list(type = "id_pattern", pattern = "p-{seq:05d}")
  )
)
out <- tempfile(fileext = ".csv")
res <- make_synthetic_csv(recipe, out)
res$rows
res$seed

df <- utils::read.csv(out)
dim(df)
names(df)
head(df, 3)
```

The seed makes the stand-in reproducible; `n_rows` overrides the recipe's own count. A
`<path>.synthetic` sidecar file is written beside the CSV (the file itself stays a plain
table), and the capsule checks read that marker, so a synthetic run can never be mistaken
for a result:

```{r synthetic-mark}
make_synthetic_csv(recipe, tempfile(fileext = ".csv"), n_rows = 5)$rows
```

`make_synthetic_column()` generates one column from a spec, for building a table piece by
piece or for testing a generator.

## Putting them together

1. On the first, trusted download: `infer_schema()`, then add the rules the schema cannot
   express (ordering between columns, cross-table consistency) and pin both.
2. On every later run: `validate_schema()` and `validate_rules()`; stop on `fatal`.
3. When the source is unreachable: `make_synthetic_csv()` from the pinned schema, run the
   pipeline, and let the synthetic marker say what the output is.
