---
title: "Year-over-year change, and the three ways it goes wrong"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Year-over-year change, and the three ways it goes wrong}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(rmoriebricklayer)
```

A year-over-year percent change is the most-quoted number in open-data
reporting and the least qualified. `yoy()` computes it, and refuses to
compute it in the three cases where the figure would describe something
other than the data.

## The data

Ontario's inmate data is published one row per placement, by fiscal year
and by group. A realistic shape:

```{r}
seg <- data.frame(
  EndFiscalYear = rep(2019:2023, each = 2),
  Gender = rep(c("Female", "Male"), 5),
  Number_Of_Placements = c(31, 402, 28, 377, 12, 190, 19, 268, 24, 331)
)
```

Segregation placements rising is bad news, so the direction is declared.
That affects colour and the verdict column, never the arithmetic.

```{r}
y <- yoy(seg,
  value = Number_Of_Placements,
  period = EndFiscalYear,
  by = "Gender",
  direction = "lower_is_better"
)
y
```

The interval is exact. Conditional on the two periods' total, the current
count is binomial, so the ratio of the two has a Clopper-Pearson
interval, which is the same construction `stats::poisson.test()` uses:

```{r}
stats::poisson.test(c(331, 268), c(1, 1))$conf.int
```

## The first failure: matching on row order

With a year missing, comparing row *i* with row *i - 1* compares 2023
with 2021 and labels it a one-year change. Matching on the period's own
value cannot do that:

```{r}
gap <- data.frame(year = c(2019, 2020, 2022, 2023), n = c(100, 120, 140, 150))
yoy(gap, value = n, period = year)
```

2021 appears as a row with no value, and 2022 has no comparison period
rather than a quietly three-year one.

## The second failure: a percent off a small base

Two placements becoming twenty is a 900% rise and also nothing at all.
The number describes how small the denominator was:

```{r}
tiny <- data.frame(year = 2019:2021, n = c(2, 20, 25))
yoy(tiny, value = n, period = year)
```

The change of `+18` is still reported, and so is the direction: those are
facts. Only the percent is withheld, with the reason attached. Lower the
gate if the small base is the point:

```{r}
as.data.frame(yoy(tiny, value = n, period = year, min_base = 0))$pct_change
```

## The third failure: a percent of a percent

If a rate moves from 4% to 5% that is one percentage point. Calling it
25% answers a different question:

```{r}
rate <- data.frame(year = 2019:2023, share = c(4.1, 4.6, 5.2, 5.0, 5.4))
yoy(rate, value = share, period = year, units = "percent")
```

The column is named `points`, and no count interval is offered, because
these are not counts.

## Seasonal series

A monthly series compares with the same month a year earlier. The lag
follows the series' own frequency, so January is never put against
December:

```{r}
m <- stats::ts(c(10:21, 20:31), start = c(2021, 1), frequency = 12)
head(as.data.frame(yoy(m, min_base = 0))[12:14, c("period", "value",
                                                  "previous", "change")], 3)
```

## Across the span

```{r}
yoy_summary(y)
```

`cagr_pct` is the compound rate per period: applying it across the span
returns `total_pct` exactly.

## Output

The same object writes to six formats. The format follows the file name.

```{r}
dir <- tempdir()
for (ext in c("csv", "tsv", "json", "md", "html", "pdf")) {
  f <- file.path(dir, paste0("placements.", ext))
  yoy_write(y, f, title = "Placements by gender")
  cat(sprintf("%-5s %6d bytes\n", ext, file.size(f)))
}
```

The delimited writers lead with the settings as comments, and carry the
`flag` column, so a withheld percentage stays marked as withheld
wherever it lands:

```{r}
cat(yoy_csv(y, NULL, digits = 1L))
```

Markdown, for a report:

```{r}
cat(yoy_markdown(yoy(gap, value = n, period = year), NULL))
```

The HTML is a single self-contained file: it makes no network request, so
the page renders later exactly as it rendered when the capsule was
sealed. It defines its palette as custom properties and carries a dark
variant. The PDF is drawn on R's own device and paginates rather than
truncating, so neither needs a package beyond base R.

```{r}
yoy_palettes()
```

`"safe"` is a blue/orange pair that survives both common forms of colour
blindness; `"mono"` emits no colour at all, for print.

```{r, include = FALSE}
unlink(file.path(dir, paste0("placements.",
                             c("csv", "tsv", "json", "md", "html", "pdf"))))
```
